---
title: "More dplyr demo"
author: "Wing-Ho Ko"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load the packages we'll need

```{r}
library(dplyr)
library(car)
library(nycflights23)
```

### Summary statistics

Keep a simpler dataframe for the summary statistics examples below

```{r}
flights_sim <- flights |> 
  select(carrier:flight | origin:dest | dep_delay | arr_delay)
```

Calculate basic summary statistics using `summary()`

```{r}
summary(flights_sim)
```

Calculate the average departure delay directly. Notice that in this case NA propagates....

```{r}
mean(flights_sim$dep_delay)
```

To omit NA's, add the named argument `na.rm=TRUE`

```{r}
mean(flights_sim$dep_delay, na.rm=TRUE)
```

Similarly, to obtain standard deviation we may do (note that R computes the standard deviation that derives from the unbiased point estimator for variance):

```{r}
sd(flights_sim$dep_delay, na.rm=TRUE)
```

For median, use `median()` instead of `mean()`

```{r}
median(flights_sim$dep_delay, na.rm=TRUE)
```

More generally, use `quantile()` for quantiles:

```{r}
quantile(flights_sim$dep_delay, 0.25, na.rm=TRUE)
```

We can bundle multiple calculations of summary statistics using `summarize()`

```{r}
flights_sim |> summarise(
  mean_delay = mean(dep_delay, na.rm=TRUE),
  sd_delay = sd(dep_delay, na.rm=TRUE)
)
```

### Grouped summary statistics

Calculate the monthly average in departure delay

```{r}
grouped <- flights |> group_by(year, month) |>
  summarize(
    dep_mean = mean(dep_delay, na.rm=TRUE),
    dep_sd = sd(dep_delay, na.rm=TRUE),
  ) |> ungroup()
grouped
```
