---
title: "Data types demo"
author: "Wing-Ho Ko"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Loading packages

```{r}
library(readr)
library(dplyr)
library(stringr)
library(forcats)
library(lubridate)
library(nycflights23)
```

### COnversion between data types

```{r}
glimpse(flights)
```

Example: convert hour and minute into integer

```{r}
flights_mod <- mutate(flights, hour=as.integer(hour), minute=as.integer(minute))
glimpse(flights_mod)
```

Example: Make origin a factor column instead of a character column

```{r}
flights_mod <- mutate(flights_mod, origin = as.factor(origin))
glimpse(flights_mod)
```

### String processing via `stringr`

Convert strings to lower case and then back

```{r}
flights_mod <- mutate(flights_mod, dest = str_to_lower(dest))
glimpse(flights_mod)
```

```{r}
flights_mod <- mutate(flights_mod, dest = str_to_upper(dest))
glimpse(flights_mod)
```

Combine origin and destination via `str_c()`

```{r}
flights_mod <- mutate(flights_mod, trip = str_c(origin, "-", dest))
glimpse(flights_mod)
```

String interpolation via `str_glue()`

```{r}
flights_mod <- mutate(
  flights_mod, 
  trip_verbose = str_glue("from {origin} to {dest}")
)
glimpse(flights_mod)
```

String subsetting via `str_sub()`

```{r}
flights_mod <- mutate(flights_mod, trip_start = str_sub(trip, 1, 3))
glimpse(flights_mod)
```

String replacement via `str_replace()`:

```{r}
flights_mod <- mutate(
  flights_mod, 
  trip_start = str_replace(trip_start, fixed("EWR"), "Newark")
)
glimpse(flights_mod)
```

### Factors processing

Glimpse at the `gss_cat` dataset included with forcats

```{r}
glimpse(gss_cat)
```

Focus on religious affiliation (relig). Obtain the list of levels using `levels()`

```{r}
levels(gss_cat$relig)
```

Use `summary()` to obtain the count for each level

```{r}
summary(gss_cat$relig)
```

Combine the "No answer", "Not applicable", and "Don't know" levels to a single "NA" level

```{r}
relig_re <- fct_recode(
  gss_cat$relig,
  "NA" = "No answer",
  "NA" = "Not applicable",
  "NA" = "Don't know"
)
```

```{r}
summary(relig_re)
```

Combine groups with fewer than 50 people into other

```{r}
relig_re <- fct_lump_min(relig_re, 50)
```

```{r}
summary(relig_re)
```

Retain only the top 10 groups, put the rest into the "Other" category

```{r}
relig_re <- fct_lump_n(relig_re, 10)
```

```{r}
summary(relig_re)
```

Reorder groups by decreasing number of respondents

```{r}
relig_re <- fct_infreq(relig_re)
```

```{r}
summary(relig_re)
```

Reorder by alphabetical order

```{r}
relig_re <- fct_relevel(relig_re, sort)
```

```{r}
summary(relig_re)
```

Put "Protestant" before "NA" so that "NA", "None", and "Other" are last

```{r}
relig_re <- fct_relevel(relig_re, "Protestant", after=6)
```

```{r}
summary(relig_re)
```

### Date and datetime processing via `lubridate`

#### Date processing via `lubridate`

Load the birthday dataset (which is modified from <https://www.kaggle.com/datasets/ulrikthygepedersen/birthdays/data>)

```{r}
bdays <- read_csv("data/birthdays.csv")
```

```{r}
glimpse(bdays)
```

Use different function to parse the 3 date representations:

```{r}
bdays_mod <- bdays |> mutate(
  date_1 = dmy(date_1),
  date_2 = mdy(date_2),
  date_3 = ymd(date_3)
)
```

```{r}
glimpse(bdays_mod)
```

Obtain the birthdays from individual component instead:

```{r}
bdays_mod <- bdays_mod |> mutate(
  date_4 = make_date(year, month, day)
)
```

```{r}
glimpse(bdays_mod)
```

Conversely, extract day of year information from date object:

```{r}
bdays_mod <- bdays_mod |> mutate(
  day_of_year = yday(date_4)
)
```

```{r}
tail(bdays_mod)
```

#### Datetime processing via `lubridate`

Load the `flights` dataset again

```{r}
glimpse(flights)
```

Use `make_datetime()` to create a datetime representation

```{r}
flights_mod <- flights |> mutate(
  datetime = make_datetime(year, month, day, hour, minute)
)
glimpse(flights_mod)
```
Alternatively, assemble a datetime string and use `ymd_hm()`:

```{r}
flights_mod <- flights_mod |> mutate(
  datetime2 = str_c(year, "-", month, "-", day, " ", hour, ":", minute)
)
glimpse(flights_mod)
```

```{r}
flights_mod <- flights_mod |> mutate(
  datetime2 = ymd_hm(datetime2)
)
glimpse(flights_mod)
```

Extract the day of year information:

```{r}
flights_mod <- flights_mod |> mutate(
  day_of_year = yday(datetime)
)
glimpse(flights_mod)
```

Check that different days of year are present

```{r}
flights_mod |> select(day_of_year) |> distinct()
```

#### Datetime arithmetic

Construct time difference

```{r}
tdiff <- as.Date("2025-01-05") - as.Date("2024-06-30")
print(tdiff)
```

Get the time difference as number of days

```{r}
tdiff / ddays(1)
```

Add the time difference to the later date

```{r}
as.Date("2025-01-05") + as.duration(tdiff)
```

Repeat the above, but with 2 times the duration

```{r}
as.Date("2025-01-05") + as.duration(tdiff) * 2
```

#### Timezone

List the names of different timezones:

```{r}
tail(OlsonNames(), 20)
```

Set the timezone using tz argument

```{r}
t1 <- as.POSIXct("2014-05-16 3:24:00", tz="US/Pacific")
t2 <- as.POSIXct("2014-05-16 3:24:00", tz="US/Eastern")
```

Time difference works as intended:

```{r}
t1 - t2
```

Convert t2 to Pacific time

```{r}
with_tz(t2, tzone = "US/Pacific")
```

Force the timezone of a datetime (changing the timezone without changing the nominal datetime value)

```{r}
force_tz(t2, tzone = "US/Pacific")
```
